### 2024-10-10 manuscript.tex
Update 06.a-gene-module-perspective-for-genetic-studies.md [ci skip]
This build is based on
https://github.com/pivlab/annual_review_of_biomedical_data_science/commit/c213a22b1a038348f3a47b835a50d5d5e35f8e7b.

This commit was created by the following CI build and job:
https://github.com/pivlab/annual_review_of_biomedical_data_science/commit/c213a22b1a038348f3a47b835a50d5d5e35f8e7b/checks
https://github.com/pivlab/annual_review_of_biomedical_data_science/actions/runs/11280147789

This manuscript
(\href{https://pivlab.github.io/annual_review_of_biomedical_data_science/v/66a49f4f63bfcb89ba3bd7ab5a0a44478728a6ec/}{permalink})
was automatically generated
from \href{https://github.com/pivlab/annual_review_of_biomedical_data_science/tree/66a49f4f63bfcb89ba3bd7ab5a0a44478728a6ec}{pivlab/annual\_review\_of\_biomedical\_data\_science@66a49f4}
on October 10, 2024.

@@

Epigenome-wide association studies (EWAS) encompass methodologies such methylome-wide association studies (MWAS) (\textbf{Figure \ref{fig:fig1}}) and histone acetylome-wide association studies (HAWAS) (\textbf{Figure \ref{fig:fig1}}), based on haQTLs and mQLTs respectability, which focus on specific types of epigenetic modifications to reveal their roles in gene regulation and disease etiology {[}\protect\hyperlink{ref-TPqvAJda}{31}{]}.
MWAS targets DNA methylation patterns, identifying loci where methylation changes are linked to particular traits or diseases.
For instance, a study by Shen et al.~demonstrated a causal relationship between DNA methylation and depression, indicating that epigenetic modifications may mediate genetic risk for psychiatric disorders {[}\protect\hyperlink{ref-1DNxNkLsT}{32}{]}.
HAWAS focuses on histone acetylation modifications, which are crucial for regulating chromatin structure and gene expression.
Del Rosario et al.conducted a HAWAS identifying over 2,000 differentially acetylated loci in immune cells from Mycobacterium tuberculosis-infected individuals, linking these changes to gene expression and potassium channel genes like \emph{KCNJ15} in modulating apoptosis and Mtb clearance.
haQTL analysis further revealed variants associated with immune phenotypes, complementing GWAS findings and enhancing understanding of disease mechanisms {[}\protect\hyperlink{ref-1BYHeZlgB}{33}{]}.
The tissue-specific nature of epigenetic modifications, alongside the capacity to capture cellular plasticity and environmental influences, enhances our insight into the effects of genetic variants across distinct tissues, temporal cellular states, and gene-environment interactions {[}\protect\hyperlink{ref-FeaSnwhX}{34},\protect\hyperlink{ref-KfWCLspO}{35}{]}.
Despite the advancements facilitated by single variant and single gene methodologies, a common thread persists: the focus remains on one gene at a time.
The underlying expectation is that identifying a gene linked to a trait will directly unveil the biological mechanisms driving disease processes.
While this has been successful in certain monogenic disorders, complex traits often involve intricate interactions among multiple genes and environmental factors.
Consequently, the one-gene-at-a-time paradigm may oversimplify the multifaceted nature of these traits.
In light of these considerations, it becomes imperative to reevaluate our approaches to dissecting the genetic architecture of complex traits.
This necessitates a shift towards methodologies that can capture the polygenic and network-based interactions inherent in complex diseases.

@@

In the early 20th century, a significant debate arose between Mendelian geneticists and biometricians concerning the inheritance of complex traits.
Mendelians emphasized single gene traits with discrete inheritance patterns, while biometricians argued that such models could not account for the continuous phenotypic variation observed in populations.
This conflict highlighted a fundamental question: how can discrete genetic factors produce continuous phenotypic variation? {[}\protect\hyperlink{ref-vpIDZCSa}{36},\protect\hyperlink{ref-15ldVppuv}{37}{]}.
The disagreement was resolved by R.A. Fisher in 1918, who introduced the infinitesimal model, demonstrating that if many genes influence a trait, their combined effects can generate continuous phenotypic distributions {[}\protect\hyperlink{ref-Qxk70FFV}{38}{]}.
Despite the success of Fisher's infinitesimal model, the actual number of genes involved in complex traits and the magnitude of their effects remained uncertain for much of the 20th century.
Translating genotypes into phenotypes proved challenging due to the complexity of the genetic mechanisms driving trait variation.
In this context, GWAS brought a new perspective for identifying specific genetic variants with functional effects on complex traits.
Over a decade of GWAS has led to several unexpected findings:
The strongest GWAS associations exhibit modest effect sizes on disease risk, collectively accounting for only a minor fraction of the predicted heritability (missing heritability) {[}\protect\hyperlink{ref-RASeYPIy}{39}{]}.
It is now understood that much of this missing heritability arises from a vast number of common variants with small effects that remain undetected in current sample sizes {[}\protect\hyperlink{ref-1AOy1zxAv}{40}{]}.
A surprising uniform distribution of these small-effect significant variants exists throughout the genome, including non-coding regions, with a notable concentration in regulatory elements such as enhancers and promoters {[}\protect\hyperlink{ref-mwTa2RUK}{41},\protect\hyperlink{ref-lVJVFaaZ}{42}{]}.
For instance, it is estimated that 71\%--100\% of 1-MB windows in the genome contribute to heritability for schizophrenia {[}\protect\hyperlink{ref-XvQe1H3A}{43}{]}.
Trait variation is influenced by the broader genome, not only by genes with direct biological links to specific phenotypes.
However, the idea that nearly all genomic variants could exert direct additive effects on a given phenotype is biologically unlikely {[}\protect\hyperlink{ref-vpIDZCSa}{36}{]}.
In response to these unexpected findings, Boyle et al.~proposed the omnigenic model {[}\protect\hyperlink{ref-vpIDZCSa}{36}{]}, a conceptual framework in which the genetic architecture of complex traits could be explained by highly interconnected gene regulatory networks.
The omnigenic model reframes the earlier infinitesimal model by distinguishing between ``core'' genes, which have a direct biological role in the disease or trait, and ``peripheral'' genes, which influence core genes and thus indirectly affect the disease or trait through regulatory networks {[}\protect\hyperlink{ref-vpIDZCSa}{36}{]}.
The omnigenic model, unlike the traditional infinitesimal model, offers a mechanistic rationale for complex traits architecture based on molecular and cellular biology, understanding the molecular pathways that connect genetic variation to phenotypic traits (\textbf{Figure \ref{fig:fig2}}).

@@

One major concern is the lack of clear, quantitative methods for identifying peripheral genes or estimating their contribution to heritability.
While the model suggests that most genetic effects arise from peripheral genes, it remains unclear how to statistically define and measure these effects {[}{[}\protect\hyperlink{ref-82fs3SZo}{48}{]}{]}{[}\protect\hyperlink{ref-doi:10.20900ux2fjpbs.20170014}{\textbf{doi:10.20900/jpbs.20170014?}}{]}.
Interestingly, research suggests that most heritability in complex traits is controlled by \emph{trans} effects rather than \emph{cis} effects, despite their small effect sizes {[}\protect\hyperlink{ref-LXvTZzEA}{44}{]}.
This paradox can be explained by natural selection acting most strongly against variants with large effects, limiting the contribution of directly biologically important genes to heritability {[}\protect\hyperlink{ref-LXvTZzEA}{44}{]}.
Thus, understanding \emph{trans} effects, where a variant at one locus influences genes at distant loci is crucial for identifying and quantifying peripheral genes and their contribution to core traits.
However, detecting and quantifying \emph{trans} effects remains a significant challenge due to their weak individual effects.
The omnigenic model postulates that nearly all genes expressed in cells relevant to a given trait have the potential to influence the regulation of core genes.
However, Yengo et al., in their recent GWAS on height, challenge this view.
They introduced the first ``saturated'' GWAS, where further increases in sample size are unlikely to yield additional genetic insights unless participant diversity or variant inclusion is expanded.
Their findings revealed that only a subset of genomic regions (\textasciitilde20\%) seem to contribute to height determination in individuals of European ancestry.
This suggests that, at least for height, not all genes play a role in the trait as predicted by the omnigenic model {[}\protect\hyperlink{ref-HFRaLwGQ}{49}{]}.
Furthermore, the omnigenic model has been criticized for their binary classification of genes into core and peripheral, which might oversimplify biological systems and potentially underestimate their true complexity {[}\protect\hyperlink{ref-82fs3SZo}{48}{]}.
It also fails to account for gene-environment interactions, which play a crucial role in shaping complex traits {[}\protect\hyperlink{ref-dg9nKuy0}{50}{]}.
Additionally, while the model provides a conceptual framework, it remains unclear how to translate it into a practical statistical model {[}\protect\hyperlink{ref-1GlQtJ9Yz}{51}{]}.
As George E. P. Box remarked, ``all models are wrong, but some are useful.'' While we acknowledge that the omnigenic model simplifies the inherent complexity of biological systems and may not universally apply to all traits, we maintain that it remains a valuable framework for elucidating genetic architectures.
The model effectively bridges quantitative and molecular genetics offering comprehensive mechanistic insights with predictions on quantitative variation.
Moreover, the omnigenic model has been instrumental in shifting the focus from single gene analyses to network-based models.
Recognizing the necessity of translating this model into practical applications, machine learning-derived gene modules, which use this concept to infer gene-gene networks, offer a promising way to ground these theoretical ideas into practical approaches.

\hypertarget{from-gene-networks-to-machine-learning-derived-gene-modules-hands-on-strategies-for-inferring-gene-gene-interactions}{%
\section{From gene networks to machine learning derived gene modules: hands-on strategies for inferring gene-gene interactions}\label{from-gene-networks-to-machine-learning-derived-gene-modules-hands-on-strategies-for-inferring-gene-gene-interactions}}

\begin{figure}
\hypertarget{fig:fig3}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/fig3.svg}
\caption{\textbf{Computational methods for machine learning derived gene module detection.}.
\textbf{a)}.
Real biological network, where nodes represent genes and edges denote the relationships between them.
It captures all possible patterns that can link genes.
Some of these relationships are gene co-expression patterns, which transcriptomics can capture, such as (1), (3) and (4).
Other patterns, such as protein-protein interactions (2), are not captured by transcriptomics and are thus not reflected in the co-expression analysis.
\textbf{b).} Different approaches for capturing machine learning-derived gene modules.
In this example, transcriptome data is used as input, but other data types such as methylomics, proteomics, and metabolomics can also be applied.
The first row illustrates the PLIER method, which uses non-negative matrix factorization and prior knowledge to align gene modules with known biological pathways.
This approach generates interpretable gene modules but only captures linear relationships.
The second row shows a vanilla VAE, which can capture both linear and nonlinear patterns but does not integrate prior knowledge, resulting in less interpretable gene modules.
The third row highlights interpretable VAEs, which incorporate prior knowledge into its decoder, achieving a balance between capturing complex relationships and maintaining biological interpretability.}\label{fig:fig3}
}
\end{figure}

\hypertarget{from-single-gene-analysis-to-gene-module-discovery-with-machine-learning-approaches}{%
\subsection{From single gene analysis to gene module discovery with machine learning approaches}\label{from-single-gene-analysis-to-gene-module-discovery-with-machine-learning-approaches}}

High-throughput technologies, particularly genome-wide gene expression profiling tools such as RNA sequencing (RNA-seq), have fundamentally transformed the landscape of molecular biology.
Unlike traditional gene-by-gene approaches, RNA-seq offers a comprehensive view of the transcriptomic landscape {[}\protect\hyperlink{ref-Zm8nV5AC}{52}{]}.
These advancements, alongside conceptual frameworks like the omnigenic model, have facilitated a paradigm shift from a single gene perspective to a module-based approach, wherein groups of genes rather than individual genes are essential for elucidating the complexities of biological networks {[}\protect\hyperlink{ref-SD886fuE}{53}{]}.

Gene modules, a key type of biological network, consist of nodes (genes) and edges that reflect coexpression relationships between them (\textbf{Figure \ref{fig:fig3}a)}).
These modules, formed by genes with coordinated expression patterns under specific biological conditions, not only reveal coexpression but could also provide valuable insights into regulatory mechanisms, gene functions, and the pathways involved in traits and diseases {[}\protect\hyperlink{ref-QUI8V4RY}{54}{]}.

Numerous approaches have been proposed for module detection in gene expression data.
Machine learning (ML) techniques are particularly proficient at addressing common challenges in omics data, such as data sparsity and high dimensionality {[}\protect\hyperlink{ref-K4KeL4gw}{55}{]}.
ML methods can compress correlations into lower-dimensional representations, facilitating the processing of large datasets to identify intricate coexpression patterns.
Additionally, ML approaches are capable of integrating various types of omics data enhancing the detection of gene modules, providing a more comprehensive understanding of biological networks {[}\protect\hyperlink{ref-K4KeL4gw}{55}{]}.

In this context, unsupervised ML approaches have gained significant popularity {[}\protect\hyperlink{ref-Ko8gMp8P}{56}{]}.
These methods facilitate the discovery of patterns and structures within the data without requiring prior knowledge of the system, allowing for the inference of novel findings and hypotheses.
This capability is particularly advantageous for gene module inference, which aims to identify groups of coexpressed and potentially co-regulated genes that contribute to similar biological functions {[}\protect\hyperlink{ref-1BVbSrr6M}{57}{]}.

The most common approach in unsupervised ML is clustering, with methods such as WGCNA, which has been widely used since the first gene expression datasets became available and remains one of the most popular tools today {[}\protect\hyperlink{ref-1BVbSrr6M}{57}{]}.
However, these methodologies typically rely on traditional correlation measures (e.g., the Pearson correlation coefficient), which are limited to capturing linear associations between continuous data.
Newer approaches, such as the Clustermatch Correlation Coefficient (CCC), have been developed to quantify both linear and non-linear correlations in complex gene expression datasets, providing a more flexible means of detecting intricate relationships that are not apparent through linear models alone {[}\protect\hyperlink{ref-YgLIM2bZ}{58}{]}.

In this section, given their successful applications, we will focus on decomposition approaches like Principal Component Analysis (PCA), and deep learning methods for inferring gene modules, highlighting their effectiveness in uncovering complex gene interactions and enhancing our understanding of biological networks (\textbf{Figure \ref{fig:fig3}b)}).
However, we acknowledge that utilizing transcriptional data alone is insufficient for fully comprehending the intricate system of biological networks; it serves as an initial step toward a more comprehensive understanding.

@@

Decomposition methods reduce complex, high-dimensional gene expression data into simpler components, revealing underlying patterns and structures that aid in interpreting large-scale datasets.
These methods outperform other unsupervised machine learning approaches in recovering gene modules by effectively capturing local co-expression effects present in only a subset of biological samples, while also allowing genes to be assigned to multiple modules {[}\protect\hyperlink{ref-1BVbSrr6M}{57}{]}.

Simple models like PCA have traditionally been used to reduce dimensionality in gene expression data, offering a basic understanding of co-expressed genes {[}\protect\hyperlink{ref-1703M7bru}{59}{]}.
However, these models often lack the ability to incorporate prior biological knowledge, which limits their interpretability and biological relevance.
To overcome these limitations, more sophisticated semi-supervised approaches such as the Pathway-Level Information Extractor (PLIER) {[}\protect\hyperlink{ref-Ki2ij7zE}{60}{]}, or GenomicSuperSignature {[}\protect\hyperlink{ref-X4fhSCkz}{61}{]} have been developed.
These methodologies combine unsupervised approaches (decomposition methodologies) with supervised approaches to infer gene modules and annotate them with prior biological knowledge.

GenomicSuperSignature employs PCA to identify gene modules applying prior knowledge for biological annotation after module discovery.
This flexibility allows it to adapt to novel datasets and provides computational efficiency, even with large datasets.
However, the lack of prior knowledge integration during training can result in modules that are less biologically interpretable compared to those from PLIER-based methods {[}\protect\hyperlink{ref-X4fhSCkz}{61}{]}.

The Pathway-Level Information Extractor (PLIER) (\textbf{Figure \ref{fig:fig3}b}) is a semi-supervised framework based on non-negative matrix factorization.
During training, it reduces high-dimensional gene expression data into a smaller set of latent variables (LVs) while aligning some of them to known pathway or gene set annotations.
These LVs are designed to capture the greatest variance in the data, where a fraction of the LVs represent known mechanisms (i.e., aligned with prior knowledge/pathways) while others capture potential technical artifacts or novel patterns.
By incorporating prior pathway/gene set information, PLIER generates interpretable latent representations.
Importantly, PLIER retains some gene modules that are not aligned with pathways, allowing researchers to distinguish LVs that represent either known patterns, technical artifacts or potential novel biological insights {[}\protect\hyperlink{ref-dg9nKuy0}{50}{]}.

PLIER has been used to integrate large-scale gene expression data using transfer learning across multiple cell types and conditions {[}\protect\hyperlink{ref-14rnBunuZ}{62}{]}.
By leveraging large datasets like recount2 {[}\protect\hyperlink{ref-GM90STke}{63}{]}, it enhances the accuracy of inferred gene modules by identifying patterns consistent across diverse data.
This application addresses the limitation of small datasets, such as those for rare diseases, by training a PLIER model on a large public data compendium and applying it to smaller datasets, resulting in models that align well with biological factors.
This approach improves the interpretability and reliability of gene expression patterns, particularly in complex experimental contexts.
PLIER can be used in specific cohort contexts, focusing on dissecting traits within smaller, more homogenous datasets.
Nandi et al.~(2024) demonstrated its utility on the Human Trisome Project (HTP) {[}\protect\hyperlink{ref-v5DtVejF}{64}/{]}, identifying gene modules that mediate the effect of karyotype on body mass index in Down syndrome, highlighting key regulators such as \emph{GPX1} and \emph{MCL1}.
This allows it to capture condition-specific regulatory mechanisms, but at the expense of broader generalizability, making it a powerful tool for understanding unique biological processes within specialized groups.
PLIER, initially designed for human gene expression data, has been successfully adapted for use in other species, demonstrating its versatility across different biological contexts.
By applying the framework to model organisms, researchers can infer gene modules in species beyond humans, expanding the utility of PLIER for comparative studies and translational research.
One such adaptation is MousiPLIER, which tailors the PLIER framework for use in mouse models, a widely used system in biomedical research {[}\protect\hyperlink{ref-1CDQw6VDO}{65}{]}.

>>>>

This manuscript
(\href{https://pivlab.github.io/annual_review_of_biomedical_data_science/v/c213a22b1a038348f3a47b835a50d5d5e35f8e7b/}{permalink})
was automatically generated
from \href{https://github.com/pivlab/annual_review_of_biomedical_data_science/tree/c213a22b1a038348f3a47b835a50d5d5e35f8e7b}{pivlab/annual\_review\_of\_biomedical\_data\_science@c213a22}
on October 10, 2024.

@@

Epigenome-wide association studies (EWAS) encompass methodologies such methylome-wide association studies (MWAS) (\textbf{Figure \ref{fig:fig1}}) and histone acetylome-wide association studies (HAWAS) (\textbf{Figure \ref{fig:fig1}}), based on haQTLs and mQLTs respectability, which focus on specific types of epigenetic modifications to reveal their roles in gene regulation and disease etiology {[}\protect\hyperlink{ref-TPqvAJda}{31}{]}.
MWAS targets DNA methylation patterns, identifying loci where methylation changes are linked to particular traits or diseases.
For instance, a study by Shen et al.~demonstrated a causal relationship between DNA methylation and depression, indicating that epigenetic modifications may mediate genetic risk for psychiatric disorders {[}\protect\hyperlink{ref-1DNxNkLsT}{32}{]}.
HAWAS focuses on histone acetylation modifications, which are crucial for regulating chromatin structure and gene expression.
Del Rosario et al.conducted a HAWAS identifying over 2,000 differentially acetylated loci in immune cells from Mycobacterium tuberculosis-infected individuals, linking these changes to gene expression and potassium channel genes like \emph{KCNJ15} in modulating apoptosis and Mtb clearance.
haQTL analysis further revealed variants associated with immune phenotypes, complementing GWAS findings and enhancing understanding of disease mechanisms {[}\protect\hyperlink{ref-1BYHeZlgB}{33}{]}.
The tissue-specific nature of epigenetic modifications, alongside the capacity to capture cellular plasticity and environmental influences, enhances our insight into the effects of genetic variants across distinct tissues, temporal cellular states, and gene-environment interactions {[}\protect\hyperlink{ref-FeaSnwhX}{34},\protect\hyperlink{ref-KfWCLspO}{35}{]}.

Despite the advancements facilitated by single variant and single gene methodologies, a common thread persists: the focus remains on one gene at a time.
The underlying expectation is that identifying a gene linked to a trait will directly unveil the biological mechanisms driving disease processes.
While this has been successful in certain monogenic disorders, complex traits often involve intricate interactions among multiple genes and environmental factors.
Consequently, the one-gene-at-a-time paradigm may oversimplify the multifaceted nature of these traits.
In light of these considerations, it becomes imperative to reevaluate our approaches to dissecting the genetic architecture of complex traits.
This necessitates a shift towards methodologies that can capture the polygenic and network-based interactions inherent in complex diseases.

@@

In the early 20th century, a significant debate arose between Mendelian geneticists and biometricians concerning the inheritance of complex traits.
Mendelians emphasized single gene traits with discrete inheritance patterns, while biometricians argued that such models could not account for the continuous phenotypic variation observed in populations.
This conflict highlighted a fundamental question: how can discrete genetic factors produce continuous phenotypic variation? {[}\protect\hyperlink{ref-vpIDZCSa}{36},\protect\hyperlink{ref-15ldVppuv}{37}{]}.
The disagreement was resolved by R.A. Fisher in 1918, who introduced the infinitesimal model, demonstrating that if many genes influence a trait, their combined effects can generate continuous phenotypic distributions {[}\protect\hyperlink{ref-Qxk70FFV}{38}{]}.
Despite the success of Fisher's infinitesimal model, the actual number of genes involved in complex traits and the magnitude of their effects remained uncertain for much of the 20th century.
Translating genotypes into phenotypes proved challenging due to the complexity of the genetic mechanisms driving trait variation.
In this context, GWAS brought a new perspective for identifying specific genetic variants with functional effects on complex traits.
Over a decade of GWAS has led to several unexpected findings:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  The strongest GWAS associations exhibit modest effect sizes on disease risk, collectively accounting for only a minor fraction of the predicted heritability (missing heritability) {[}\protect\hyperlink{ref-RASeYPIy}{39}{]}.
  It is now understood that much of this missing heritability arises from a vast number of common variants with small effects that remain undetected in current sample sizes {[}\protect\hyperlink{ref-1AOy1zxAv}{40}{]}.
\item
  A surprising uniform distribution of these small-effect significant variants exists throughout the genome, including non-coding regions, with a notable concentration in regulatory elements such as enhancers and promoters {[}\protect\hyperlink{ref-mwTa2RUK}{41},\protect\hyperlink{ref-lVJVFaaZ}{42}{]}.
  For instance, it is estimated that 71\%--100\% of 1-MB windows in the genome contribute to heritability for schizophrenia {[}\protect\hyperlink{ref-XvQe1H3A}{43}{]}.
\item
  Trait variation is influenced by the broader genome, not only by genes with direct biological links to specific phenotypes.
  However, the idea that nearly all genomic variants could exert direct additive effects on a given phenotype is biologically unlikely {[}\protect\hyperlink{ref-vpIDZCSa}{36}{]}.
\end{enumerate}

In response to these unexpected findings, Boyle et al.~proposed the omnigenic model {[}\protect\hyperlink{ref-vpIDZCSa}{36}{]}, a conceptual framework in which the genetic architecture of complex traits could be explained by highly interconnected gene regulatory networks.
The omnigenic model reframes the earlier infinitesimal model by distinguishing between ``core'' genes, which have a direct biological role in the disease or trait, and ``peripheral'' genes, which influence core genes and thus indirectly affect the disease or trait through regulatory networks {[}\protect\hyperlink{ref-vpIDZCSa}{36}{]}.
The omnigenic model, unlike the traditional infinitesimal model, offers a mechanistic rationale for complex traits architecture based on molecular and cellular biology, understanding the molecular pathways that connect genetic variation to phenotypic traits (\textbf{Figure \ref{fig:fig2}}).

@@

One major concern is the lack of clear, quantitative methods for identifying peripheral genes or estimating their contribution to heritability.
While the model suggests that most genetic effects arise from peripheral genes, it remains unclear how to statistically define and measure these effects {[}\protect\hyperlink{ref-82fs3SZo}{48},\protect\hyperlink{ref-javtAG7r}{49}{]}.
Interestingly, research suggests that most heritability in complex traits is controlled by \emph{trans} effects rather than \emph{cis} effects, despite their small effect sizes {[}\protect\hyperlink{ref-LXvTZzEA}{44}{]}.
This paradox can be explained by natural selection acting most strongly against variants with large effects, limiting the contribution of directly biologically important genes to heritability {[}\protect\hyperlink{ref-LXvTZzEA}{44}{]}.
Thus, understanding \emph{trans} effects, where a variant at one locus influences genes at distant loci is crucial for identifying and quantifying peripheral genes and their contribution to core traits.
However, detecting and quantifying \emph{trans} effects remains a significant challenge due to their weak individual effects.

The omnigenic model postulates that nearly all genes expressed in cells relevant to a given trait have the potential to influence the regulation of core genes.
However, Yengo et al., in their recent GWAS on height, challenge this view.
They introduced the first ``saturated'' GWAS, where further increases in sample size are unlikely to yield additional genetic insights unless participant diversity or variant inclusion is expanded.
Their findings revealed that only a subset of genomic regions (\textasciitilde20\%) seem to contribute to height determination in individuals of European ancestry.
This suggests that, at least for height, not all genes play a role in the trait as predicted by the omnigenic model {[}\protect\hyperlink{ref-HFRaLwGQ}{50}{]}.

Furthermore, the omnigenic model has been criticized for their binary classification of genes into core and peripheral, which might oversimplify biological systems and potentially underestimate their true complexity {[}\protect\hyperlink{ref-82fs3SZo}{48}{]}.
It also fails to account for gene-environment interactions, which play a crucial role in shaping complex traits {[}\protect\hyperlink{ref-dg9nKuy0}{51}{]}.
Additionally, while the model provides a conceptual framework, it remains unclear how to translate it into a practical statistical model {[}\protect\hyperlink{ref-1GlQtJ9Yz}{52}{]}.

As George E. P. Box remarked, ``all models are wrong, but some are useful.'' While we acknowledge that the omnigenic model simplifies the inherent complexity of biological systems and may not universally apply to all traits, we maintain that it remains a valuable framework for elucidating genetic architectures.
The model effectively bridges quantitative and molecular genetics offering comprehensive mechanistic insights with predictions on quantitative variation.
Moreover, the omnigenic model has been instrumental in shifting the focus from single gene analyses to network-based models.
Recognizing the necessity of translating this model into practical applications, machine learning-derived gene modules, which use this concept to infer gene-gene networks, offer a promising way to ground these theoretical ideas into practical approaches.

\hypertarget{from-gene-networks-to-machine-learning-derived-gene-modules-hands-on-strategies-for-inferring-gene-gene-interactions}{%
\subsection{From gene networks to machine learning derived gene modules: hands-on strategies for inferring gene-gene interactions}\label{from-gene-networks-to-machine-learning-derived-gene-modules-hands-on-strategies-for-inferring-gene-gene-interactions}}

\begin{figure}
\hypertarget{fig:fig3}{%
\centering
\includegraphics[width=1\textwidth,height=\textheight]{images/fig3.svg}
\caption{\textbf{Computational methods for machine learning derived gene module detection.}.
\emph{(a)} Real biological network, where nodes represent genes and edges denote the relationships between them.
It captures all possible patterns that can link genes.
Some of these relationships are gene co-expression patterns, which transcriptomics can capture, such as (1), (3) and (4).
Other patterns, such as protein-protein interactions (2), are not captured by transcriptomics and are thus not reflected in the co-expression analysis.
\emph{(b)} Different approaches for capturing machine learning-derived gene modules.
In this example, transcriptome data is used as input, but other data types such as methylomics, proteomics, and metabolomics can also be applied.
The first row illustrates the PLIER method, which uses non-negative matrix factorization and prior knowledge to align gene modules with known biological pathways.
This approach generates interpretable gene modules but only captures linear relationships.
The second row shows a vanilla VAE, which can capture both linear and nonlinear patterns but does not integrate prior knowledge, resulting in less interpretable gene modules.
The third row highlights interpretable VAEs, which incorporate prior knowledge into its decoder, achieving a balance between capturing complex relationships and maintaining biological interpretability.}\label{fig:fig3}
}
\end{figure}

\hypertarget{from-single-gene-analysis-to-gene-module-discovery-with-machine-learning-approaches}{%
\subsubsection{From single gene analysis to gene module discovery with machine learning approaches}\label{from-single-gene-analysis-to-gene-module-discovery-with-machine-learning-approaches}}

High-throughput technologies, particularly genome-wide gene expression profiling tools such as RNA sequencing (RNA-seq), have fundamentally transformed the landscape of molecular biology.
Unlike traditional gene-by-gene approaches, RNA-seq offers a comprehensive view of the transcriptomic landscape {[}\protect\hyperlink{ref-Zm8nV5AC}{53}{]}.
These advancements, alongside conceptual frameworks like the omnigenic model, have facilitated a paradigm shift from a single gene perspective to a module-based approach, wherein groups of genes rather than individual genes are essential for elucidating the complexities of biological networks {[}\protect\hyperlink{ref-SD886fuE}{54}{]}.

Gene modules, a key type of biological network, consist of nodes (genes) and edges that reflect coexpression relationships between them (\textbf{Figure \ref{fig:fig3}\emph{a}}).
These modules, formed by genes with coordinated expression patterns under specific biological conditions, not only reveal coexpression but could also provide valuable insights into regulatory mechanisms, gene functions, and the pathways involved in traits and diseases {[}\protect\hyperlink{ref-QUI8V4RY}{55}{]}.

Numerous approaches have been proposed for module detection in gene expression data.
Machine learning (ML) techniques are particularly proficient at addressing common challenges in omics data, such as data sparsity and high dimensionality {[}\protect\hyperlink{ref-K4KeL4gw}{56}{]}.
ML methods can compress correlations into lower-dimensional representations, facilitating the processing of large datasets to identify intricate coexpression patterns.
Additionally, ML approaches are capable of integrating various types of omics data enhancing the detection of gene modules, providing a more comprehensive understanding of biological networks {[}\protect\hyperlink{ref-K4KeL4gw}{56}{]}.

In this context, unsupervised ML approaches have gained significant popularity {[}\protect\hyperlink{ref-Ko8gMp8P}{57}{]}.
These methods facilitate the discovery of patterns and structures within the data without requiring prior knowledge of the system, allowing for the inference of novel findings and hypotheses.
This capability is particularly advantageous for gene module inference, which aims to identify groups of coexpressed and potentially co-regulated genes that contribute to similar biological functions {[}\protect\hyperlink{ref-1BVbSrr6M}{58}{]}.

The most common approach in unsupervised ML is clustering, with methods such as WGCNA, which has been widely used since the first gene expression datasets became available and remains one of the most popular tools today {[}\protect\hyperlink{ref-1BVbSrr6M}{58}{]}.
However, these methodologies typically rely on traditional correlation measures (e.g., the Pearson correlation coefficient), which are limited to capturing linear associations between continuous data.
Newer approaches, such as the Clustermatch Correlation Coefficient (CCC), have been developed to quantify both linear and non-linear correlations in complex gene expression datasets, providing a more flexible means of detecting intricate relationships that are not apparent through linear models alone {[}\protect\hyperlink{ref-YgLIM2bZ}{59}{]}.

In this section, given their successful applications, we will focus on decomposition approaches like Principal Component Analysis (PCA), and deep learning methods for inferring gene modules, highlighting their effectiveness in uncovering complex gene interactions and enhancing our understanding of biological networks (\textbf{Figure \ref{fig:fig3}\emph{b}}).
However, we acknowledge that utilizing transcriptional data alone is insufficient for fully comprehending the intricate system of biological networks; it serves as an initial step toward a more comprehensive understanding.

@@

Decomposition methods reduce complex, high-dimensional gene expression data into simpler components, revealing underlying patterns and structures that aid in interpreting large-scale datasets.
These methods outperform other unsupervised machine learning approaches in recovering gene modules by effectively capturing local co-expression effects present in only a subset of biological samples, while also allowing genes to be assigned to multiple modules {[}\protect\hyperlink{ref-1BVbSrr6M}{58}{]}.

Simple models like PCA have traditionally been used to reduce dimensionality in gene expression data, offering a basic understanding of co-expressed genes {[}\protect\hyperlink{ref-1703M7bru}{60}{]}.
However, these models often lack the ability to incorporate prior biological knowledge, which limits their interpretability and biological relevance.
To overcome these limitations, more sophisticated semi-supervised approaches such as the Pathway-Level Information Extractor (PLIER) {[}\protect\hyperlink{ref-Ki2ij7zE}{61}{]}, or GenomicSuperSignature {[}\protect\hyperlink{ref-X4fhSCkz}{62}{]} have been developed.
These methodologies combine unsupervised approaches (decomposition methodologies) with supervised approaches to infer gene modules and annotate them with prior biological knowledge.

GenomicSuperSignature employs PCA to identify gene modules applying prior knowledge for biological annotation after module discovery.
This flexibility allows it to adapt to novel datasets and provides computational efficiency, even with large datasets.
However, the lack of prior knowledge integration during training can result in modules that are less biologically interpretable compared to those from PLIER-based methods {[}\protect\hyperlink{ref-X4fhSCkz}{62}{]}.

The Pathway-Level Information Extractor (PLIER) (\textbf{Figure \ref{fig:fig3}\emph{b}}) is a semi-supervised framework based on non-negative matrix factorization.
During training, it reduces high-dimensional gene expression data into a smaller set of latent variables (LVs) while aligning some of them to known pathway or gene set annotations.
These LVs are designed to capture the greatest variance in the data, where a fraction of the LVs represent known mechanisms (i.e., aligned with prior knowledge/pathways) while others capture potential technical artifacts or novel patterns.
By incorporating prior pathway/gene set information, PLIER generates interpretable latent representations.
Importantly, PLIER retains some gene modules that are not aligned with pathways, allowing researchers to distinguish LVs that represent either known patterns, technical artifacts or potential novel biological insights {[}\protect\hyperlink{ref-dg9nKuy0}{51}{]}.

PLIER has been used to integrate large-scale gene expression data using transfer learning across multiple cell types and conditions {[}\protect\hyperlink{ref-14rnBunuZ}{63}{]}.
By leveraging large datasets like recount2 {[}\protect\hyperlink{ref-6SPTvFXq}{64}{]}, it enhances the accuracy of inferred gene modules by identifying patterns consistent across diverse data.
This application addresses the limitation of small datasets, such as those for rare diseases, by training a PLIER model on a large public data compendium and applying it to smaller datasets, resulting in models that align well with biological factors.
This approach improves the interpretability and reliability of gene expression patterns, particularly in complex experimental contexts.

PLIER can be used in specific cohort contexts, focusing on dissecting traits within smaller, more homogenous datasets.
Nandi et al.~(2024) demonstrated its utility on the Human Trisome Project (HTP) {[}\protect\hyperlink{ref-v5DtVejF}{65}{]}, identifying gene modules that mediate the effect of karyotype on body mass index in Down syndrome, highlighting key regulators such as \emph{GPX1} and \emph{MCL1}.
This allows it to capture condition-specific regulatory mechanisms, but at the expense of broader generalizability, making it a powerful tool for understanding unique biological processes within specialized groups.

PLIER, initially designed for human gene expression data, has been successfully adapted for use in other species, demonstrating its versatility across different biological contexts.
By applying the framework to model organisms, researchers can infer gene modules in species beyond humans, expanding the utility of PLIER for comparative studies and translational research.
One such adaptation is MousiPLIER, which tailors the PLIER framework for use in mouse models, a widely used system in biomedical research {[}\protect\hyperlink{ref-1CDQw6VDO}{66}{]}.
